---
title: "Getting Started with wARMASVp"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with wARMASVp}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6, fig.height = 4
)
```

The **wARMASVp** package provides closed-form estimation, simulation, hypothesis
testing, and forecasting for higher-order stochastic volatility SV(p) models.
It supports Gaussian, Student-t, and Generalized Error Distribution (GED)
innovations, with optional leverage effects.

## The SV(p) Model

The stochastic volatility model of order $p$ is:

$$y_t = \sigma_y \exp(w_t / 2)\, z_t$$
$$w_t = \phi_1 w_{t-1} + \cdots + \phi_p w_{t-p} + \sigma_v v_t$$

where $z_t$ is an i.i.d. innovation (Gaussian, Student-t, or GED) and
$v_t \sim N(0,1)$ drives the log-volatility.

## Simulation and Estimation

### Gaussian SV(1)

```{r gaussian-sv1}
library(wARMASVp)
set.seed(123)

# Simulate
y <- sim_svp(2000, phi = 0.95, sigy = 1, sigv = 0.3)

# Estimate
fit <- svp(y, p = 1, J = 10)
summary(fit)
```

### Gaussian SV(2)

```{r gaussian-sv2}
y2 <- sim_svp(2000, phi = c(0.20, 0.63), sigy = 1, sigv = 0.5)
fit2 <- svp(y2, p = 2, J = 10)
summary(fit2)
```

### Student-t Innovations

```{r student-t}
yt <- sim_svp(2000, phi = 0.90, sigy = 1, sigv = 0.3,
              errorType = "Student-t", nu = 5)
fit_t <- svp(yt, p = 1, errorType = "Student-t")
summary(fit_t)
```

### GED Innovations

```{r ged}
yg <- sim_svp(2000, phi = 0.90, sigy = 1, sigv = 0.3,
              errorType = "GED", nu = 1.5)
fit_ged <- svp(yg, p = 1, errorType = "GED")
summary(fit_ged)
```

## Leverage Effects

When return and volatility shocks are correlated ($\rho \neq 0$), use the
leverage option:

```{r leverage}
sim_lev <- sim_svp(2000, phi = 0.95, sigy = 1, sigv = 0.3,
                   leverage = TRUE, rho = -0.5)
fit_lev <- svp(sim_lev$y, p = 1, leverage = TRUE)
summary(fit_lev)
```

## Hypothesis Testing

The package provides Local Monte Carlo (LMC) and Maximized Monte Carlo (MMC)
tests. Here we test for AR order (is SV(1) sufficient vs SV(2)?):

```{r test-ar}
# Generate from SV(1)
y_test <- sim_svp(2000, phi = 0.95, sigy = 1, sigv = 0.3)

# Test H0: SV(1) vs H1: SV(2) — should not reject
test_ar <- lmc_ar(y_test, p_null = 1, p_alt = 2, N = 49)
print(test_ar)
```

Test for leverage:

```{r test-lev}
test_lev <- lmc_lev(y_test, p = 1, N = 49)
print(test_lev)
```

## Forecasting

Multi-step ahead volatility forecasts using Kalman filtering:

```{r forecast}
y_fc <- sim_svp(500, phi = 0.95, sigy = 1, sigv = 0.3,
                leverage = TRUE, rho = -0.3)$y
fc <- forecast_svp(y_fc, p = 1, H = 20)
plot(fc)
```

## References

- Ahsan, N. and Dufour, J.-M. (2021). Simple estimators and inference for
  higher-order stochastic volatility models. *Journal of Econometrics*,
  224(1), 181-197.
- Ahsan, N., Dufour, J.-M., and Rodriguez Rondon, G. (2025). Estimation and
  inference for higher-order stochastic volatility models with leverage.
  *Journal of Time Series Analysis*.
